A Bayesian networks approach to infer social changes in burials in northeastern Taiwan during the European colonization period

### 2021-06-09 analysis/paper/paper.Rmd
finished editing for reviews

# Exponential random graph models in a Bayesian framework {-}

@@

# Background {-}

## A Case Study from Northeastern Taiwan {-}

<!--Intro to our case study and model-->
We use a Bayesian ERGM to study social changes in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan (Figure \@ref(fig:figure-KWL-map)), which was occupied from the 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Spanish and the Dutch in the 17th century, and finally an influx of Han Chinese in the 19th century [@Chen2007]. Our anthropological model proposes that the influence of an indirect colonial power, combined with high local values attached to imported goods, led to increased social inequality due to competition among individuals [@Brumfiel1994; @Clark1994]. Before the arrival of Europeans, some imported goods already entered Kiwulan because of the regional trade network with Chinese traders [@Chen2005; @Cheng2017]. We assume that the use of imported goods and their values were amplified after the European arrival. Different from Heping dao and Tamsui in northern Taiwan where European forts are located, the remote geographical setting of northeastern Taiwan made it less interfered by European colonial power but experienced frequent trade activities due to local natural resources, such as rice and deerskins [@Chen2005; @Cheng2017]. This makes northeastern Taiwan a special area to explore the effect of European colonization in a pericolonial context that is still not well understood compared to European colonial bases [@Acabado2017; @Trabert2017].

<!--theoretical framework: corporate/network strategies-->
The involvement in long-distance exchange is often associated with changes in political-economic strategies from a corporate mode to a network mode, where ambitious individuals are able to build their local power base and personal influence through the acquisition and distribution of foreign goods [@Blanton1996; @Feinman2000; @Klehm2017]. The features of network-based societies include the accumulation of prestige goods, wealth differentiation, and trade monopolization through individual networks [@Blanton1996; @Feinman2000]. These are contrasted with corporate-based societies that stress communal ritual elements, shared power, and where wealth differentiation, if any, would be associated with corporate groups, such as age-based or kin-based, rather than individuals [@Siegel1999]. Network-based societies are identified in some places in Asia that were involved in long-distance trade [@Carter2015; @Ueda2016; @Liu2006].

@@

<!--hypothesis and research questions-->
Using the theoretical framework of corporate-network strategies, we test a hypothesis of the shift in organizational modes in relation to increased social inequality, from a more corporate to a more network society, after the European presence that can be observed in the Kiwulan burial network. We ask: (1) Do the burial network after European arrival present a more centralized structure with fewer subgroups compared to the burial network before European contact? and if so, (2) what are the major variables affecting or forming these two networks? Can we identify the shift in variables from ritual to wealth associated with two different strategies? By answering these questions, this case study helps to expand our understanding of European colonial effects on Indigenous groups in a pericolonial context.

@@

## Materials {-}

@@

<!-- Our chronology and principles of network building -->
To compare burial networks, we assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`). Our assignments are based on an established fine-grained chronology that was reexamined and cross-validated by diagnostic materials, stratigraphic data, depth, and radiocarbon ages [for more details see @LWandBM2020ornament; @LWandBM2020pot]. We excluded burials from the Chinese phase (n = `r chi`) due to the smaller sample size. We also excluded `r disturb` burials that were heavily disturbed by modern construction which prevented accurate determination of their chronology. Differences in network size were evaluated using vertex bootstrap in a later section.

Based on the assumption that social status and associated relations can be represented by sharing similar prestige goods [cf. @Coward2013, pp. 252], we built networks where burials (nodes in the network) are linked when they have the same trade goods in common. The trade goods we identified include gold-foil beads, carnelian beads, glass beads, Chinese porcelains, stonewares, gold foils, fish-shaped ornaments, and imported coins, which were also less affected by taphonomic processes because of durable materials. These items are considered as prestige goods due to their rarity in burial contexts and descriptions in historical records [@Chen2007; @Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011]. European historical accounts mentioned that those items were treated as prestige goods in Indigenous culture [@Borao2009; @LiandWu2006]. For example, a Spanish priest described a Spanish soldier exchanging carnelian beads for natural resources with local Indigenous people, because of the beads’ high value in Indigenous culture [@LiandWu2006]. In addition, Spanish visitors observed that Indigenous people possessing more imported goods were recognized as having higher status in their community [@LiandWu2006].

@@

```{r bergm-pre, include=FALSE}
suppressWarnings(source(here::here("analysis", "code", "003-burials-pre-network-modelling.R")))
```

```{r bergm-post, include=FALSE}
suppressWarnings(source(here::here("analysis", "code", "004-burials-post-network-modelling.R")))
```

## Hypothesis and construction of networks {-}

<!-- our expectation based on hypothesis --> 
Our social change hypothesis is that differential access to foreign prestige goods after the European presence led to increased social inequality at Kiwulan, where the social structure changed from a more corporate strategy to a more network strategy, presenting different organizational modes of societies [@Blanton1996; @Drennan2010; @Feinman2000]. Corporate-based societies stress shared power, multiple segments, and social cohesion through communal rituals, while network-based societies emphasize the accumulation of individual prestige, wealth, and resources by building personal networks. We used burial data as a proxy to explore social changes from a corporate strategy to a network strategy. This change can be identified though the pattern of intercorrelations of foreign prestige goods based on assumptions that burials can represent social structures [@Binford1971; @Saxe1970], and individuals with similar prestige goods reflect similar access to trade, exchange, and gifting networks [eg. @Coward2013, pp. 252].

If social inequality gradually increased after European arrival as we hypothesize, then we expect to observe a network structure with higher centralization and less transitivity, along with network ties determined by wealth, indicative of a network-based society. In contrast, we predict the network before European arrival will show low centralization and high transitivity, and network ties associated with ritual elements, indicative of a corporate-based society. High transitivity (or clustering) indicates multiple subgroups with shared burial goods, representing social cohesion instead of individual accumulation in an earlier time. To test this hypothesis, we use Bayesian ERGMs to model the formation of network ties and the underlying mechanisms that shaped relationships between people at Kiwulan. By comparing networks from the pre-European period and the post-European period we can examine the effects of foreign contact on community relationships at Kiwulan.

@@

<!-- burial data and the definition of tie formation  --> 
Trade beads are commonly found across burials with substantial differences in quantities, so we described each burial as having one of four levels, high, upper-middle, lower-middle, and low, according to their distributions across all burials (Figure \@ref(fig:raincloud-beads); details in Supplementary Online Materials). For less frequent prestige goods, including imported ceramics, gold foils, and fish-shaped ornaments, we linked two burials when they both possess each type of goods (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and a burial value index. Ritual pottery was identified as locally-made ceramics placed above an individual. These pots mostly have sooty vessel exteriors and charred residues on interiors that are interpreted as vessels used for funeral feasting [@Hsieh2009]. We calculated our burial value by summing values of all types of prestige goods in each burial context. For example, for gold-foil beads, we take the total number of burials (90) and divide by number of burials with gold-foil beads (46), to give a value of 90/46 = 1.96. We repeat this for each type of prestige item. For each burial we sum these values of each type of prestige item to get a burial value [@Jorgensen1991].

@@

We then assigned the burials into three ranks according to breaks in the distribution of burial values, high (>`r high_value`, top 10 percent), above average (`r ave_value`-`r high_value`), and below average (<`r ave_value`), as an index of wealth (details in Supplementary Online Materials). Since burials tend to have multiple goods in common, the network ties are weighted instead of binary (the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means a mutual relationship where the tie between any two actors is bidirectional. The networks constructed based on these principles show that the network after the European presence has more node connectivity in general with some nodes having a larger number of connections (Figure \@ref(fig:figure-network-diagrams)).

```{r figure-network-diagrams, fig.cap = "A: Burial network before the European arrival (14th-16th centuries), B: Burial network after the European arrival (17th-18th centuries). The size of each node is proportional to node degrees or the number of connections to a node. The thickness of ties represents the number of goods in common between two nodes. Isolated nodes are not displayed in the plots"}
knitr::include_graphics(here::here("analysis", "figures", "005-network-diagrams.png"))
```

@@

<!-- Network construction and specification -->
We quantify the relations among burials and test our hypothesis of social change using the R programming language [@Rlanguage2019] with the bergm package [@Caimo2014]. Table \@ref(tab:table-ergm-terms) lists the network parameters we used for dependence assumptions that define our models [@Morris2008]. Every parameter in an ERGM has an associated algorithm for computing the probability of observing relations between two burials. Based on our hypothesis, we model a network with increased social inequality to be represented by endogenous network effects: low transitivity and high centralization. We include burial-specific attributes as covariate effects for homophily, such as age, sex, ritual activity, and the degree of wealth, to explore the importance of these variables in tie forming. For example, if age-homophily is important here, then people of the same age should have the same burial goods. We also include the physical distance between burials as an indicator of a kinship-based relations since the deceased from the same family were buried nearby [@LiandWu2006]. Our model may reveal the emergence of social inequality via the presence of a few individuals as network centers, having more relations with others, with more wealth differentiation after the European arrival.

@@

To evaluate our anthropological model of increased social inequality over time based on assumptions of corporate-network strategies [@Drennan2010; @Feinman2000], we specify priors that are meaningful for social inequality, especially for the variables of transitivity and centralization. We set the priors to higher transitivity (N(`r priors$pre_prior_mean[6]`, `r priors$pre_prior_sigma[6]`)), lower centralization (N(`r priors$pre_prior_mean[7]`, `r priors$pre_prior_sigma[7]`)), and higher covariate effect of ritual activity (N(`r priors$pre_prior_mean[4]`, `r priors$pre_prior_sigma[4]`)) for the network before European contact to indicate less social differentiation and emphasize the ritual element shared in corporate groups. Conversely, we set the priors for the network after European arrival to lower transitivity (N(`r priors$post_prior_mean[6]`, `r priors$post_prior_sigma[6]`)), higher centralization (N(`r priors$post_prior_mean[7]`, `r priors$post_prior_sigma[7]`)), and a higher covariate effect for burial values (N(`r priors$post_prior_mean[5]`, `r priors$post_prior_sigma[5]`)) to model an increased social differentiation after European contact. For covariates about biological features, such as age, sex, we specified a vague prior (N(`r priors$mean_same[2]`, `r priors$sigma_same[2]`)) for both models to explore their effects. For physical distance between burials, we also set a vague prior (N(`r priors$mean_same[8]`, `r priors$sigma_same[8]`)) to explore whether there is kinship-based proximity, e.g. stronger correlations for shorter distances.

@@

<!--posterior stats for pre-E-->
We examine the estimates from the posterior distributions to compare their differences in structure of the simulated networks (Table \@ref(tab:table-posterior-stats); Figure \@ref(fig:figure-posterior-distribution)). For nodal covariates in the pre-European model, the values of ritual element represented by pots (ritual-homophily) are mostly greater than zero that suggests a positive effect on the formation of relations between burials, while the values of wealth rank represented by burial values (wealth-homophily) are mostly negative that indicates it has less influence on the tie formation. This demonstrates that burials with ritual pottery tend to form relations, but burials in the same wealth level tend not to. For other covariates with values mostly around zero, such as age (age-homophily) and sex (sex-homophily), they do not show any clear tendency. Similarly, the dyadic covariate, physical distance, shows no significant effect, indicating that physical proximity between burials does not reflect similarity in burial treatment. For the endogenous network effects, transitivity presents a significant positive effect, while centralization demonstrates a negative effect. The high positive value for transitivity suggests a tendency of burials with similar burial goods to be clustered as connected communities, indicative of the presence of multiple corporate groups sharing burial goods in common. In contrast, the strong negative centralization shows there is a tendency toward decentralization that reflects most burials having a similar number of ties without any prominent burials. This might imply that individuals have equal access to trade goods in terms of the flow of goods.

@@

library(flextable)
# add column labels
ft <- flextable(posterior_esti_two_phases) %>% 
  colformat_double(digits = 2) %>% 
  set_header_labels(mean_pre = "pre-European", median_pre = "pre-European", 
                    `2.5%_pre` = "pre-European", `97.5%_pre` = "pre-European",
                    mean_post = "post-European", median_post = "post-European", 
                    `2.5%_post` = "post-European", `97.5%_post` = "post-European") %>%
  merge_at(i = 1, j = 2:5, part = "header") %>% 
  merge_at(i = 1, j = 6:9, part = "header") %>% 
  add_header_row(values = c("", "mean", "median", "2.5%", "97.5%",
                            "mean", "median", "2.5%", "97.5%"), top = FALSE ) %>% 
  align(j = 2:9, align = "right", part = "header") %>% 
  align(i = 1, align = "center", part = "header") %>% 
  width(j = ~ parameter, width = 1.36) %>% 
  width(j = 2:9, width = 0.67) %>% 
  set_caption("Estimated posterior means, medians, and 95% confidence intervals for each network parameter of two models. Confidence intervals that do not include zero indicate significant effects of parameters (Caimo, 2017). Posterior means are generally close to posterior medians.")
ft
```

```{r figure-posterior-distribution, fig.cap = "Posterior density estimates for the parameters associated with edges, transitivity, centralization, and physical distance by phases. The pre-European group presents remarkably larger values for the transitivity, but lower values for the edges and centralization parameters compared to the post-European group. The distance parameters overlap between two groups. Both posterior estimates present generally symmetric distributions."}
knitr::include_graphics(here::here("analysis", "figures", "006-posterior-distribution.png"))
```

<!--posterior stats for for post-E and their comparison-->
For the post-European network model, two nodal covariates, wealth rank and age, both show tendencies towards positive values. This indicates the burials in the same wealth level and age groups tend to form relations. However, there is no clear pattern for ritual and sex variables since most values around zero. Also, there are no significant effects for the variables of physical distance. Similar to the pre-European network, transitivity demonstrates a significant positive effect, but weaker than the effect of the pre-European network. In contrast, centralization has a significantly higher positive effect than the effect of the pre-European network. This means there is a tendency toward centralization, reflecting a limited number of burials having many more ties than others. This implies that the presence of better access to trade goods and the behavior of wealth accumulation and display in burial events. In general, the post-European network model has a smaller transitivity effect and a positive centralization effect than the pre-European network model. This may suggest a reduced tendency toward clustering but high tendency toward centralization after the European presence.

```{r bootstrap-CI-test, include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/code/007-bootstrap-CI.R"))}
suppressWarnings(source(here::here("analysis", "code", "007-bootstrap-CI.R")))
```

<!-- comparison between two networks using bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with `r pre_E` burials compared to `r post_E` burials. To understand the robustness of comparison between two networks, we used a vertex bootstrap technique to cross-validate the results of our Bayesian ERGMs. The vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. node) to quantify standard errors and estimate sampling variability in the network statistics of interest [@Chen2019; @Snijders1999; @Rroberts2021]. This enables the evaluation of uncertainty for networks and tests the difference between multiple networks of different sizes by examining their confidence intervals for the network population. We used the vertex bootstrap to compute endogenous network statistics, including density, centralization, and transitivity for our two networks. We explored the sample size effect by removing nodes at certain percentages (5-40%) for both networks and comparing their confidence intervals. The results demonstrate a consistent difference between the two networks up to 30% node removals as shown in Figure \@ref(fig:figure-bootstrap-CI). This indicates that the network variables, including density, centralization, and transitivity, are robust to the different network size of our sample. Figure \@ref(fig:figure-bootstrap-CI) also shows consistency with our earlier finding of negative centralization in the pre-European period and positive centralization in the post-European period, over several conditions of node removal. Similarly consistent with our earlier result, the vertex bootstrap shows no significant difference on density and transitivity.

```{r figure-bootstrap-CI, fig.cap = "Results of the vertex bootstrap analysis by three network variables, showing the 95% bootstrap confidence intervals for the bootstrapped (n = 1000) networks, for each category of node removal. The 0% node removal category is the original data for the two phases. Networks with nonoverlapping confidence intervals indicate a significant difference between the two phases for that network variable."}
knitr::include_graphics(here::here("analysis", "figures", "007-bootstrap-CI.png"))
```

@@

```{r figure-distribution-moments, fig.cap = "Distribution moments (mean, skewness, and variance) calculated on the observed data and simulated distributions for the pre- and post-European models."}
knitr::include_graphics(here::here("analysis", "figures", "006-distribution-moments.png"))
```

@@

<!-- theoretical framework for the results-->
A striking finding in our results is the change in network properties of Kiwulan burials from a more cohesive network with multiple subgroups toward a more centralized network with concentration of connections among fewer burials after the European arrival in the 17th century. This supports our pericolonial impact model that Kiwulan changed from a more corporate-based to a more networked-based society, as indicated in the grave goods and burial attributes. Changes in mortuary practices could reflect the development of social or political complexity [eg. @Curet1998]. A corporate-based society, as we interpret the pre-European situation at Kiwulan, stresses shared power between individuals, communal rituals, and social inequality, if any, would be associated with groups [@Drennan2010; @Feinman2000; @Feinman2000political]. In contrast, a network-based society, as we interpret the post-European data from Kiwulan, presents wealth accumulation through individual networks, prestige goods manipulation, and trade monopolization [@Blanton1996]. It should be noted that the corporate-network continuum represents a dynamic process with different degrees of hierarchical complexity instead of static ideal-type stages [@Feinman2000].

>>>>

## Exponential random graph models in a Bayesian framework {-}

@@

# Case Study: Exploring Social Changes in Northeastern Taiwan {-}

<!--Intro to our case study and model-->
We use a Bayesian ERGM to study social changes in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan (Figure \@ref(fig:figure-KWL-map)), which was occupied from the 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Spanish and the Dutch in the 17th century, and finally an influx of Han Chinese in the 19th century [@Chen2007]. Our anthropological model proposes that the influence of an indirect colonial power, combined with high local values attached to imported goods, led to increased social inequality due to competition among individuals (details in Supplementary Online Materials) [@Brumfiel1994; @Clark1994]. Before the arrival of Europeans, some imported goods already entered Kiwulan because of the regional trade network with Chinese traders [@Chen2005; @Cheng2017]. We assume that the use of imported goods and their values were amplified after the European arrival. Different from Heping dao and Tamsui in northern Taiwan where European forts are located, the remote geographical setting of northeastern Taiwan made it less interfered by European colonial power but experienced frequent trade activities due to local natural resources, such as rice and deerskins [@Chen2005; @Cheng2017]. This makes northeastern Taiwan a special area to explore the effect of European colonization in a pericolonial context that is still not well understood compared to European colonial bases [@Acabado2017; @Trabert2017].

<!--theoretical framework: corporate/network strategies-->
The involvement in long-distance exchange is often associated with an increased social inequality where ambitious individuals could gain power by building personal influence through the acquisition and distribution of foreign goods, leading stoical change from corporate based to network based societies [@Blanton1996; @Feinman2000; @Klehm2017]. Corporate-network societies represent two distinct political-economic strategies for obtaining power, viewed as a horizontal axis crosscutting traditional vertical hierarchy [@Drennan2010; @Feinman2000]. The features of network strategies include the accumulation of prestige goods, wealth differentiation, and trade monopolization through individual networks. These are contrasted with corporate strategies that stress communal ritual elements, shared power, and where wealth differentiation, if any, would be associated with corporate groups, such as age-based or kin-based, rather than individuals [@Siegel1999]. In Asia, network-based societies are identified, especially for places involved in long-distance trade [@Carter2015; @Ueda2016; @Liu2006].

@@

<!--hypothesis and research questions-->
Using the theoretical framework of corporate-network strategies, we test a hypothesis of the shift in organizational modes in relation to increased social inequality, from a more corporate to a more network society, after the European presence that can be observed in the Kiwulan burial network [@Blanton1996; @Drennan2010; @Feinman2000]. We explore whether this change can be identified through the pattern of intercorrelations of foreign prestige goods among burials. We ask: (1) Do the burial network after European arrival present a more centralized structure with fewer subgroups compared to the burial network before European contact? and if so, (2) what are the major variables affecting or forming these two networks? Can we identify the shift in variables from ritual to wealth associated with two different strategies? By answering these questions, this case study helps to expand our understanding of European colonial effects on Indigenous groups in a pericolonial context.

@@

# Materials {-}

@@

<!--assigning burials based on our chronology-->
To compare burial networks, we assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`). Our assignments are based on an established fine-grained chronology that was reexamined and cross-validated by diagnostic materials, stratigraphic data, depth, and radiocarbon ages [for more details see @LWandBM2020ornament; @LWandBM2020pot]. We excluded burials from the Chinese phase (n = `r chi`) due to the smaller sample size. We also excluded `r disturb` burials that were heavily disturbed by modern construction which prevented accurate determination of their chronology. Differences in network size were evaluated using vertex bootstrap in a later section. By comparing networks from the pre-European period and the post-European period we can examine the effects of foreign contact on community relationships at Kiwulan.

## Contextualizing Burial Networks {-}

<!--the context of burial networks-->
We built networks where burials (nodes in the network) are linked when they have the same trade goods in common. This is based on the assumptions that social status and associated relations can be represented by sharing similar goods and burials can reflect social structures [@Binford1971; @Saxe1970; cf. @Coward2013, pp. 252]. Thus, individuals with the same prestige goods reflect similar access to trade, exchange, and gifting networks, indicative of the similarity among Kiwulan residents and social inequality. The trade goods we identified include gold-foil beads, carnelian beads, glass beads, Chinese porcelains, stonewares, gold foils, fish-shaped ornaments, and imported coins, which were also items less affected by taphonomic processes because of durable materials. Those goods are considered as prestige goods due to their rarity in burial contexts and descriptions in historical records [@Chen2007; @Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011]. European historical accounts mentioned that those items were treated as prestige goods in Indigenous culture [@Borao2009; @LiandWu2006]. For example, a Spanish priest described a Spanish soldier exchanging carnelian beads for natural resources with local Indigenous people, because of the beads’ high value in Indigenous culture [@LiandWu2006]. In addition, Spanish visitors observed that Indigenous people possessing more imported goods were recognized as having higher status in their community [@LiandWu2006].

@@

```{r bergm-pre, include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/code/003-burials-post-network-modelling.R"))}
suppressWarnings(source(here::here("analysis", "code", "003-burials-pre-network-modelling.R")))
```

```{r bergm-post, include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/code/004-burials-post-network-modelling.R"))}
suppressWarnings(source(here::here("analysis", "code", "004-burials-post-network-modelling.R")))
```

## Hypothesis and prediction {-}

<!-- archaeological predictions --> 
Our social change hypothesis is that differential access to foreign prestige goods after the European presence led to increased social inequality at Kiwulan, where the social structure changed from a more corporate strategy to a more network strategy, presenting different organizational modes of societies [@Blanton1996; @Drennan2010; @Feinman2000]. If social inequality gradually increased after European arrival as we hypothesize, then we expect to observe a network structure with higher centralization and less transitivity, along with network ties determined by wealth, indicative of a network-based society. In contrast, we predict the network before European arrival will show low centralization and high transitivity, and network ties associated with ritual elements, indicative of a corporate-based society. High transitivity (or clustering) indicates multiple subgroups with shared burial goods, representing social cohesion instead of individual accumulation in an earlier time. To test this hypothesis, we use Bayesian ERGMs to model the formation of network ties and the underlying mechanisms that shaped relationships between people at Kiwulan.

@@

## Construction of Network Variables {-}

<!-- burial data and the definition of tie formation  --> 
Trade beads are found across burials with substantial differences in quantities, so we described each burial as having one of four levels, high, upper-middle, lower-middle, and low, according to their distributions across all burials (Figure \@ref(fig:raincloud-beads); details in Supplementary Online Materials). For less frequent prestige goods, including imported ceramics, gold foils, fish-shaped ornaments, and coins, we linked two burials when they both possess each type of goods (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and a burial value index. Ritual pottery was identified as locally-made ceramics placed above an individual, which is interpreted as vessels used for funeral feasting or offering [@Hsieh2009]. We calculated our burial value by summing values of all types of prestige goods in each burial context. For example, for gold-foil beads, we take the total number of burials (90) and divide by the number of burials with gold-foil beads (46), to give a value of 90/46 = 1.96. We repeat this for each type of prestige item. For each burial we sum these values of each type of prestige item to get a burial value [@Jorgensen1991].

@@

We then assigned the burials into three ranks according to breaks in the distribution of burial values, high (>`r high_value`, top 10 percent), above average (`r ave_value`-`r high_value`), and below average (<`r ave_value`), as an index of wealth. Since burials tend to have multiple goods in common, the network ties are weighted instead of binary (the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. We are able to weigh ties due to durable materials of trade goods that are preserved well in each burial. Our networks are non-directed, which means a mutual relationship where the tie between any two actors is bidirectional. The networks constructed based on these principles show that the network after the European presence has more node connectivity in general with some nodes having a larger number of connections (Figure \@ref(fig:figure-network-diagrams)).

```{r figure-network-diagrams, fig.cap = "A: Burial network before the European arrival (14th-16th centuries), B: Burial network after the European arrival (17th-18th centuries). Nodes represent burials, tied when they share the same trade goods. The size of each node is proportional to node degrees or the number of connections to a node. The thickness of ties represents the number of goods in common between two nodes. Isolated nodes are not displayed in the plots."}
knitr::include_graphics(here::here("analysis", "figures", "005-network-diagrams.png"))
```

@@

<!-- Network construction and specification -->
We quantify the relations among burials and test our hypothesis of social change using the R programming language [@Rlanguage2019] with the bergm package [@Caimo2014]. Table \@ref(tab:table-ergm-terms) lists the network parameters we used for dependence assumptions that define our models [@Morris2008]. Every parameter in an ERGM has an associated algorithm for computing the probability of observing relations between two burials. Based on our hypothesis, we model a network with increased social inequality to be represented by endogenous network effects: low transitivity and high centralization. We include burial-specific attributes as covariate effects for homophily, such as age, sex, ritual activity, and the degree of wealth, to explore the importance of these variables in tie forming. For example, if age-homophily is important here, then people of the same age should have the same burial goods. We also include the physical distance between burials as an indicator of a kinship-based relations that we assume the deceased from the same family were buried nearby [@Chen2007]. Our model may reveal the emergence of social inequality via the presence of a few individuals as network centers, having more relations with others, with more wealth differentiation after the European arrival.

@@

To evaluate our anthropological model of increased social inequality over time to indicate a shift from a corporate to a network society [@Drennan2010; @Feinman2000], we specify priors that are meaningful for social change, especially for the variables of transitivity and centralization. We set the priors to higher transitivity (N(`r priors$pre_prior_mean[6]`, `r priors$pre_prior_sigma[6]`)), lower centralization (N(`r priors$pre_prior_mean[7]`, `r priors$pre_prior_sigma[7]`)), and higher covariate effect of ritual activity (N(`r priors$pre_prior_mean[4]`, `r priors$pre_prior_sigma[4]`)) for the network before European contact to indicate less social differentiation and emphasize the ritual element shared in corporate groups. Conversely, we set the priors for the network after European arrival to lower transitivity (N(`r priors$post_prior_mean[6]`, `r priors$post_prior_sigma[6]`)), higher centralization (N(`r priors$post_prior_mean[7]`, `r priors$post_prior_sigma[7]`)), and a higher covariate effect for burial values (N(`r priors$post_prior_mean[5]`, `r priors$post_prior_sigma[5]`)) to model an increased social differentiation after European contact. For covariates about biological features, such as age, sex, we specified a vague prior (N(`r priors$mean_same[2]`, `r priors$sigma_same[2]`)) for both models to explore their effects. Similarly, we used the same prior (N(`r priors$mean_same[8]`, `r priors$sigma_same[8]`)) for physical distance between burials, to explore whether there is kinship-based proximity, e.g. stronger correlations for shorter distances.

@@

```{r bgof-assessment, include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/code/007-bgof-assessment-vis.R"))}
suppressWarnings(source(here::here("analysis", "code", "007-bgof-assessment-vis.R")))
```

<!--posterior stats for pre-E-->
We examine the estimates from the posterior distributions to compare their differences in structure of the simulated networks (Table \@ref(tab:table-posterior-stats); Figure \@ref(fig:figure-posterior-distribution)). For nodal covariates in the pre-European model, the values of ritual element represented by pots (ritual-homophily) are mostly greater than zero that suggests a positive effect on the formation of relations between burials, while the values of wealth rank represented by burial values (wealth-homophily) are mostly negative that indicates it has less influence on the tie formation. This demonstrates that burials with ritual pottery tend to form relations, but burials in the same wealth level tend not to. For other covariates with values mostly around zero, such as age (age-homophily) and sex (sex-homophily), they do not show any clear tendency. For the physical distance variable (dyadic covariate), it shows that most values are positive, indicating that physical proximity between burials could reflect similarity in burial treatment. For the endogenous network effects, transitivity presents a significant positive effect, while centralization demonstrates a negative effect. The high positive value for transitivity suggests a tendency of burials with similar burial goods to be clustered as connected communities, indicative of the presence of multiple corporate groups sharing burial goods in common. In contrast, the strong negative centralization shows there is a tendency toward decentralization that reflects most burials having a similar number of ties without any prominent burials. This might imply that individuals have equal access to trade goods in terms of the flow of goods.

@@

library(flextable)
# add column labels
ft <- flextable(posterior_esti_two_phases) %>% 
  colformat_double(digits = 2) %>% 
  set_header_labels(mean_pre = "pre-European", median_pre = "pre-European", 
                    `2.5%_pre` = "pre-European", `97.5%_pre` = "pre-European",
                    mean_post = "post-European", median_post = "post-European", 
                    `2.5%_post` = "post-European", `97.5%_post` = "post-European") %>%
  merge_at(i = 1, j = 2:5, part = "header") %>% 
  merge_at(i = 1, j = 6:9, part = "header") %>% 
  add_header_row(values = c("", "mean", "median", "2.5%", "97.5%",
                            "mean", "median", "2.5%", "97.5%"), top = FALSE ) %>% 
  align(j = 2:9, align = "right", part = "header") %>% 
  align(i = 1, align = "center", part = "header") %>% 
  width(j = ~ parameter, width = 1.36) %>% 
  width(j = 2:9, width = 0.67) %>% 
  set_caption("Estimated posterior means, medians, and 95% confidence intervals for each network parameter of two models. Confidence intervals that do not include zero indicate significant effects of parameters (Caimo, 2017). Some posterior means close to posterior medians that indicate symmetric distributions.")
ft
```

```{r figure-posterior-distribution, fig.cap = "Posterior density estimates for the parameters associated with edges, transitivity, centralization, and physical distance by phases. The pre-European group presents a long tailed distribution with larger values for the transitivity, but negative values for the centralization parameter compared to the post-European group. The edges and distance parameters for the pre-European group have larger values generally compared to pos-European one."}
knitr::include_graphics(here::here("analysis", "figures", "006-posterior-distribution.png"))
```

<!--posterior stats for for post-E and their comparison-->
For the post-European network model, two nodal covariates, wealth rank and age, both show tendencies towards positive values. This indicates the burials in the same wealth level and age groups tend to form relations. However, there is no clear pattern for ritual and sex variables since most values around zero. Also, there are no effects for the variable of physical distance. Similar to the pre-European network, transitivity demonstrates a significant positive effect, but weaker than the effect of the pre-European network. In contrast, centralization has a significantly higher positive effect than the effect of the pre-European network. This means there is a tendency toward centralization, reflecting a limited number of burials having many more ties than others. This implies that the presence of better access to trade goods and the behavior of wealth accumulation and display in burial events. In general, the post-European network model has a smaller transitivity effect and higher centralization effect than the pre-European network model. This may suggest a reduced tendency toward clustering but high tendency toward centralization after the European presence.

```{r bootstrap-CI-test, include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/code/008-bootstrap-CI.R"))}
suppressWarnings(source(here::here("analysis", "code", "008-bootstrap-CI.R")))
```

<!-- comparison between two networks using bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with `r pre_E` burials compared to `r post_E` burials. To understand the robustness of comparison between two networks, we used a vertex bootstrap technique to cross-validate the results of our Bayesian ERGMs. The vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. node) to quantify standard errors and estimate sampling variability in the network statistics of interest [@Chen2019; @Snijders1999; @Rroberts2021]. This enables the evaluation of uncertainty for networks and tests the difference between multiple networks of different sizes by examining their confidence intervals for the network population. We used the vertex bootstrap to compute endogenous network statistics, including density, centralization, and transitivity for our two networks. We explored the sample size effect by removing nodes at certain percentages (5-40%) for both networks and comparing their confidence intervals. The results demonstrate a consistent difference between the two networks up to 35% node removals for the centralization variable as shown in Figure \@ref(fig:figure-bootstrap-CI). This shows consistency with our earlier finding of negative centralization in the pre-European period and positive centralization in the post-European period, over several conditions of node removal. Similarly consistent with our earlier result, the vertex bootstrap shows no significant difference on density and transitivity. This indicates that the network variables, including density, centralization, and transitivity, are robust to the different network size of our sample.

```{r figure-bootstrap-CI, fig.cap = "Results of the vertex bootstrap analysis by three network variables, showing the 95% bootstrap confidence intervals for the bootstrapped (n = 1000) networks, for each category of node removal. The 0% node removal category is the original data for the two phases. Networks with nonoverlapping confidence intervals indicate a significant difference between the two phases for that network variable."}
knitr::include_graphics(here::here("analysis", "figures", "008-bootstrap-CI.png"))
```

@@

```{r figure-distribution-moments, fig.cap = "Distribution moments (mean, skewness, and variance) calculated on the observed data and simulated distributions for the pre- and post-European models."}
knitr::include_graphics(here::here("analysis", "figures", "007-distribution-moments.png"))
```

@@

<!-- theoretical framework for the results-->
A striking finding in our results is the change in network properties of Kiwulan burials from a more cohesive network with multiple subgroups toward a more centralized network with concentration of connections among fewer burials after the European arrival in the 17th century. This supports our pericolonial impact model that Kiwulan changed from a more corporate-based to a more networked-based society, as indicated in the burial goods and burial attributes. Changes in mortuary practices could reflect the development of social or political complexity [eg. @Curet1998]. A corporate-based society, as we interpret the pre-European situation at Kiwulan, stresses shared power between individuals, communal rituals, and social inequality, if any, would be associated with groups [@Drennan2010; @Feinman2000; @Feinman2000political]. In contrast, a network-based society, as we interpret the post-European data from Kiwulan, presents wealth accumulation through individual networks, prestige goods manipulation, and trade monopolization [@Blanton1996]. It should be noted that the corporate-network continuum represents a dynamic process with different degrees of hierarchical complexity instead of static ideal-type stages [@Feinman2000].
